\chapter*{Acknowledgements} %the star makes the chapter unnumbered
I would like to express my sincere gratitude to those who have supported and guided me.

First and foremost, I extend my deepest appreciation to my thesis supervisor, \textbf{Dr. Mario Hernandez Tinoco}, for his invaluable guidance, encouragement, and insightful feedback, which have been instrumental in shaping this research.

Throughout this journey, I have been fortunate to exchange ideas with peers who share a deep interest in quantitative finance and market dynamics.

I am profoundly grateful to \textbf{Wadi Moughanim} (École Normale Supérieure, 2024), Oil Quantitative Analyst at Litasco, Geneva, whose support in shaping the research topic and expertise in commodity markets significantly contributed to the foundation of this study. A special thank you to \textbf{Antoine Monestier} (Université Paris Dauphine, 2026), Equity Derivatives Trader at BBVA, Hong Kong, for his invaluable guidance in refining my approach. I sincerely appreciate \textbf{Julien Meunier} (Sciences Po Paris, 2025), Investment Banker at Morgan Stanley, Paris, for his help in reviewing my paper, as well as \textbf{Guillaume Nodet} (École Polytechnique, 2024), Quantitative Analyst at Barclays, London, for his insights into quantitative analysis that enriched this research.

I am also grateful to \textbf{Thomas Stokolm Jensen}, Investment Advisor at UBS, Geneva, for his valuable insights into investment strategies, and his help into reaching to \textbf{Hannes Huett}, Executive Director at UBS, Zurich, for Quantitative Strategies, their expertise in quantitative finance strengthened the analytical foundation of this work.

I am deeply grateful to \textbf{EDHEC Business School}, which has provided a stimulating academic environment and essential resources for my research.

In particular, I would like to thank \textbf{Professor Arnaud Dufays}, Professor of Quantitative Methods, for his crucial feedback and review of my paper. I also extend my gratitude to: Professor \textbf{Gisèle Hites}, for her insights into Machine Learning ; Professor \textbf{Simo Alami Chehboune}, for his expertise in Time Series. EDHEC also provided the \textbf{Wharton Research Data Services (WRDS)} that played a crucial role in data acquisition and analysis, greatly supporting the empirical aspects of this study.

Finally, I would like to express my deepest gratitude to my \textbf{family and friends} for their unwavering encouragement, patience, and support throughout this journey. Their belief in me has been a constant source of motivation.